\LARGE \textbf{Constrained Geometric Attitude Control on \( \SO \)}

### 2016-12-15 manuscript.tex
Rearrange the figures to flow better

\section{Numerical Examples}\label{sec:numerical_simulation}
We demonstrate the performance of the proposed control system via numerical simulation.
The inertia tensor of a rigid body is given as
\begin{gather*}
  \hspace{-5mm}
	J = \begin{bmatrix}
	\num{5.57e-3} & \num{6.17e-5} & \num{-2.50e-5} \\
	\num{6.17e-5} & \num{5.57e-3} & \num{1.00e-5} \\
	\num{-2.50e-5} & \num{1.00e-5} & \num{1.05e-2}
	\end{bmatrix} \si{\kilo\gram\meter\squared} .
\end{gather*} 
The control system parameters are chosen as
\begin{gather*}
	G = \text{diag} [0.9,1.1,1.0], \quad k_R = 0.4 , \quad	k_\Omega = 0.296 ,\\
	c = 1.0 , \quad k_\Delta = 0.5 , \quad \alpha = 15 .
\end{gather*}
The diagonal matrix \( G \) serves as a weighting matrix for the relative difference between \( R \) and \( R_d \). 
Using this term, the control designer can modify the shape of the attractive error function, given in~\cref{eqn:A}, and the resulting behavior of the closed loop system.
Similarly, the constant \( \alpha \) is used to modify the shape of the repulsive error function, given in~\cref{eqn:B}.
In general, this term is derived from the system design and the nature of the obstacles in the dynamic environment.
For example, a system wishing to avoid pointing at a diffuse obstacle, such as incoming debris, may chose an appropriate value of \( \theta \), based on the best available information, and a relatively low \( \alpha \) to ensure additional safety margin near the constraint boundary. 
Similarly, in an environment with several densely spaced obstacles, such as that presented in~\cref{fig:cad_adapt}, a much larger \( \alpha \) would enable more aggressive manuevers which pass closer to the constraint boundary without violation.
This would increase the allowable region of operation in a highly constrained environment.
The parameters \( k_R, k_\Omega, c, k_\Delta\) are control parameters used to modify the closed-loop behavior of the system.
It is straightforward to chose \( k_R, k_\Omega, k_\Delta\), using a linear analysis, to satisfy desired response criteria, such as settling time or percent overshoot~\cite{nise2004}.
A body fixed sensor is defined as \(r = [1,0,0]\), while multiple inequality constraints are defined in~\Cref{tab:constraints}.
The simulation parameters are chosen to be similar to those found in~\cite{lee2011b}, however we increase the size of the constraint regions to create a more challenging scenario for the control system.
\begin{table}[H]
\caption{Constraint Parameters~\label{tab:constraints}}
\begin{center}\begin{tabular}{lc}
Constraint Vector (\( v \)) & Angle (\( \theta \)) \\ \hline \hline 
\([0.174,\,-0.934,\, -0.034]^T\) & \ang{40} \\ \hline 
\([0 ,\, 0.7071 ,\, 0.7071]^T\) & \ang{40} \\ \hline 
\([-0.853 ,\, 0.436 ,\, -0.286]^T\) & \ang{40} \\ \hline 
\([-0.122 ,\,-0.140,\, -0.983]^T\) & \ang{20}\end{tabular} 
\end{center}
\end{table}

The initial state is defined as \(R_0 =  \exp(\ang{225} \times \frac{\pi}{180} \hat{e}_3), \Omega_0 = 0\). 
The desired state is \( R_d = I,\Omega_d = 0\).
\end{multicols}
\begin{figure}
    \centering 
    \subfigure[Attitude error vector \(e_R\) components \label{fig:eR_con}]{\includegraphics[width=0.3\columnwidth]{eR_con}  }~
    \subfigure[Configuration error \( \Psi \) \label{fig:Psi_con}]{\includegraphics[width=0.3\columnwidth]{figures/Psi_con.pdf} }
    \subfigure[Angle to each constraint \label{fig:con_angles_con}]{\includegraphics[width=0.3\columnwidth]{figures/constraint_angles_con.pdf} } 
    \caption{Attitude stabilization without adaptive update law}
    \label{fig:con} 
\end{figure}
\begin{multicols}{2}
We show simulation results for the system stabilizing about the desired attitude with and without the adaptive update law from~\Cref{prop:adaptive_control}.
We assume a fixed disturbance of \(\Delta = \begin{bmatrix} 0.2 & 0.2 & 0.2 \end{bmatrix}^T \si{\newton\meter}\), with the function \( W(R,\Omega) = I \).
This form is equivalent to an integral control term which penalizes deviations from the desired configuration.
The first term of~\cref{eqn:delta_dot} has the effect of increasing the proportional gain of the control system, since the time derivative of the attitude error vector, \( \dot{e}_{R} \), is linear with respect to the angular velocity error vector \( e_\Omega\).

@@

\subsection{Attitude Parameterizations}\label{ssec:attitude_parameterization}
% Discuss the benefits of using this geometric representation as compared to Euler angles or quaternions
Attitude parameterizations, such as Euler angles and Quaternions, are frequently used in the aerospace and astrodynamics communities~\cite{vallado2007}.
For example, Euler angle sequences are frequently used to describe the transformation between a variety of reference frames used to describe the position and orientation of the orbit of Earth satellites~\cite{vallado2007}.
In addition, quaternions were used during the operation of Skylab and the NASA Space Shuttle~\cite{hughes2004}.
However, the choice of attitude parameterization plays a critical role in control design and the resulting motion of the system.

% singularities and the problems involved with using Euler angles
Euler angle sequences are a minimum, three-parameter set of angles which describe the transformation between two reference frames.
Using Euler angles, we can represent any general rotation as a sequence of three intermediate rotations~\cite{shuster1993}.
By convention, there are \num{24} possible Euler angle sequences for any given rotation.
In addition, Euler angles are a minimum representation, as only three angles, and the associated sequence, are required to describe the three angular degrees of freedom of the rigid body.
However, there is great ambiguity in the representation of the attitude as there are many equivalent Euler angle sequences for a given attitude of the system.
Therefore, great care must be taken in the control system design to ensure that a consistent sequence is used. 
Furthermore, it has been shown that no minimal attitude representation can describe orientations both globally and without singularities~\cite{hughes2004,bhat2000}.
These singularities can cause significant difficulties during control design and hardware implementation.
\begin{figure} 
  \centering 
    \subfigure[Configuration error \( \Psi \)\label{fig:Psi_adapt}]{\includegraphics[width=0.4\columnwidth]{Psi_adapt} }~
    \subfigure[Angle to each constraint \label{fig:con_angles}]{\includegraphics[width=0.4\columnwidth]{constraint_angles} }\\
    \subfigure[Disturbance estimate \( \bar \Delta \) components \label{fig:delta_adapt} ]{\includegraphics[width=0.4\columnwidth]{Delta_adapt}}~
    \subfigure[Attitude trajectory \label{fig:cad_adapt}]{\includegraphics[trim={10cm 0 10cm 0},clip,width=0.4\columnwidth]{cad_adapt} }
  \caption{Attitude stabilization with adaptive update law}
  \label{fig:adapt} 
\end{figure}
\begin{multicols}{2}
To demonstrate the effect of the kinematic singularities inherent with Euler angles we will represent the attitude of the body fixed reference frame, \( \vecbf{b}_i \), with respect to the inertial frame, \( \vecbf{e}_i\), in terms of the \( 3-1-3\) Euler angle sequence.
More explicitly, this corresponds to the rotation sequence \( \theta_1 \vecbf{b}_3 , \theta_2 \vecbf{b}_1, \theta_3 \vecbf{b}_3 \).
The rotation matrix, \( R(\theta_1, \theta_2, \theta_3) \), corresponding to this sequence is 
\begin{align}\label{eq:euler313}
    \begin{bmatrix}
        -s_1 c_2 s_3 + c_3 c_1 & -s_1 c_2 c_3 - s_3 c_1 & s_1s_2 \\
        c_1 c_2 s_3 + c_3 s_1 & c_1 c_2 c_3 - s_3 s_1 & - c_1 s_2 \\
        s_2 s_3 & s_2 c_3 & c_2
    \end{bmatrix} ,
\end{align}
where \( s_i, c_i \) represent \( \sin \theta_i, \cos \theta_i \) for \( i = \braces{1,2,3}\).
Using this representation that kinematic differential equations for the associated Euler angles are given as
\begin{align}\label{eq:euler313_diff}
    \begin{bmatrix}
        \dot{\theta}_1 \\ \dot{\theta}_2 \\ \dot{\theta}_3 
    \end{bmatrix}
    =
    \begin{bmatrix}
        \slfrac{\parenth{\Omega_1 s_3 + \Omega_2 c_3}}{s_2} \\
        \Omega_1 c_3 - \Omega_2 s_3 \\
        -\slfrac{\parenth{\Omega_1 s_3 + \omega_2 c_3}c_2}{s_2} + \Omega_3
    \end{bmatrix} .
\end{align}
From~\cref{eq:euler313_diff}, it is immediately clear that a singularity exists when \( \sin \theta_2 = 0 \) or equivalently, \( \theta_2 = 0, \pm \pi \). 
In the vicinity of the singularity, the angular velocities of the Euler angles will tend to approach \( \pm \infty \) and the angular velocities will experience instantaneous sign changes.
Furthermore, all Euler angle sequences will exhibit a similar singularity at either \( \theta_2 = 0, \pm \pi \) or \( \theta_2 = \pm \frac{\pi}{2}, \pm \frac{3\pi}{2} \).
Therefore simply switching the sequence does not alleviate the issue, but rather only moves the singularity.
As a result, Euler angles are not appropriate for systems which experience large angular rotations, such as those demonstrated in~\cref{fig:adapt}, or control systems which rely on the angular velocities \( \theta_i \).
% discuss the unwinding phenomenon with quaternions

\subsection{Time-varying Disturbance}\label{ssec:time_varying}

@@

\Cref{fig:tv} demonstrates the ability for the adaptive controller, which is presented in~\Cref{prop:adaptive_control}, to handle time-varying disturbances.
\Cref{fig:Psi_tv} shows the non-dimensional value of the configuration error function and demonstrates that the adaptive controller is able to stabilize the system to the desired attitude configuration.
In addition,~\cref{fig:con_angle_tv} shows that the constraint is never violated as the angle between the body-fixed sensor \( r \) and the constraint \( v \) is greater than \SI{12}{\degree} over the entire attitude maneuver.
We can see in~\cref{fig:Delta_tv} that that estimate \(\bar \Delta \) for each of the components accurately tracks the true disturbance after approximately~\SI{5}{\second}.
\end{multicols}
\begin{figure}[H]
    \centering 
    \subfigure[{Configuration error \( \Psi \)} \label{fig:Psi_tv} ]{\includegraphics[width=0.3\columnwidth]{Psi_tv} }~
    \subfigure[{Angle to constraint } \label{fig:con_angle_tv}]{\includegraphics[width=0.3\columnwidth]{figures/constraint_angle_tv.pdf} }~
    \subfigure[{Disturbance estimate \(\bar\Delta\) components} \label{fig:Delta_tv}]{\includegraphics[width=0.3\columnwidth]{figures/delta_tv} }
    \caption{Time-varying external disturbance simulation}
    \label{fig:tv} 
\end{figure}
\begin{multicols}{2}
\section{Experiment on Hexrotor UAV}\label{sec:experiment}
\begin{figure}[H]
    \centering
    \includegraphics[width = 0.75\columnwidth]{hexrotor}
    \caption{Attitude control testbed~\label{fig:hexrotor}}
\end{figure}
A hexrotor unmanned aerial vehicle (UAV), as seen in~\cref{fig:hexrotor}, has been developed at the Flight Dynamics and Controls Laboratory (FDCL) at the George Washington University~\cite{kaufman2014}.
The UAV is composed of three pairs of counter-rotating propellers. 
Typical UAVs are composed of four or more co-planar propellers.
As a result, these systems are underactuated and unable to impart a force along every degree of freedom.
For example, quadrotor UAVs are unable to translate laterally without first conducting a rotation.
Conversely, the propeller pairs of the hexrotor are angled relative to one another to allow for a fully actuated rigid body.
This allows the hexrotor to impart a force in any direction and a moment about any axis.

>>>>

\end{multicols}
\begin{figure}[b]
    \centering 
    \subfigure[Attitude error vector \(e_R\) components \label{fig:eR_con}]{\includegraphics[width=0.3\columnwidth]{eR_con}  }~
    \subfigure[Configuration error \( \Psi \) \label{fig:Psi_con}]{\includegraphics[width=0.3\columnwidth]{figures/Psi_con.pdf} }
    \subfigure[Angle to each constraint \label{fig:con_angles_con}]{\includegraphics[width=0.3\columnwidth]{figures/constraint_angles_con.pdf} } 
    \caption{Attitude stabilization without adaptive update law}
    \label{fig:con} 
\end{figure}
\begin{multicols}{2}
\section{Numerical Examples}\label{sec:numerical_simulation}

We demonstrate the performance of the proposed control system via numerical simulation.
The inertia tensor of a rigid body is given as
\begin{gather*}
  \hspace{-5mm}
	J = \begin{bmatrix}
	\num{5.57e-3} & \num{6.17e-5} & \num{-2.50e-5} \\
	\num{6.17e-5} & \num{5.57e-3} & \num{1.00e-5} \\
	\num{-2.50e-5} & \num{1.00e-5} & \num{1.05e-2}
	\end{bmatrix} \si{\kilo\gram\meter\squared} .
\end{gather*} 
The control system parameters are chosen as
\begin{gather*}
	G = \text{diag} [0.9,1.1,1.0], \quad k_R = 0.4 , \quad	k_\Omega = 0.296 ,\\
	c = 1.0 , \quad k_\Delta = 0.5 , \quad \alpha = 15 .
\end{gather*}
The diagonal matrix \( G \) serves as a weighting matrix for the relative difference between \( R \) and \( R_d \). 
Using this term, the control designer can modify the shape of the attractive error function, given in~\cref{eqn:A}, and the resulting behavior of the closed loop system.
Similarly, the constant \( \alpha \) is used to modify the shape of the repulsive error function, given in~\cref{eqn:B}.
In general, this term is derived from the system design and the nature of the obstacles in the dynamic environment.
For example, a system wishing to avoid pointing at a diffuse obstacle, such as incoming debris, may chose an appropriate value of \( \theta \), based on the best available information, and a relatively low \( \alpha \) to ensure additional safety margin near the constraint boundary. 
Similarly, in an environment with several densely spaced obstacles, such as that presented in~\cref{fig:cad_adapt}, a much larger \( \alpha \) would enable more aggressive manuevers which pass closer to the constraint boundary without violation.
This would increase the allowable region of operation in a highly constrained environment.
The parameters \( k_R, k_\Omega, c, k_\Delta\) are control parameters used to modify the closed-loop behavior of the system.
It is straightforward to chose \( k_R, k_\Omega, k_\Delta\), using a linear analysis, to satisfy desired response criteria, such as settling time or percent overshoot~\cite{nise2004}.
A body fixed sensor is defined as \(r = [1,0,0]\), while multiple inequality constraints are defined in~\Cref{tab:constraints}.
The simulation parameters are chosen to be similar to those found in~\cite{lee2011b}, however we increase the size of the constraint regions to create a more challenging scenario for the control system.
\begin{table}[H]
\caption{Constraint Parameters~\label{tab:constraints}}
\begin{center}\begin{tabular}{lc}
Constraint Vector (\( v \)) & Angle (\( \theta \)) \\ \hline \hline 
\([0.174,\,-0.934,\, -0.034]^T\) & \ang{40} \\ \hline 
\([0 ,\, 0.7071 ,\, 0.7071]^T\) & \ang{40} \\ \hline 
\([-0.853 ,\, 0.436 ,\, -0.286]^T\) & \ang{40} \\ \hline 
\([-0.122 ,\,-0.140,\, -0.983]^T\) & \ang{20}\end{tabular} 
\end{center}
\end{table}

The initial state is defined as \(R_0 =  \exp(\ang{225} \times \frac{\pi}{180} \hat{e}_3), \Omega_0 = 0\). 
The desired state is \( R_d = I,\Omega_d = 0\).
We show simulation results for the system stabilizing about the desired attitude with and without the adaptive update law from~\Cref{prop:adaptive_control}.
We assume a fixed disturbance of \(\Delta = \begin{bmatrix} 0.2 & 0.2 & 0.2 \end{bmatrix}^T \si{\newton\meter}\), with the function \( W(R,\Omega) = I \).
This form is equivalent to an integral control term which penalizes deviations from the desired configuration.
The first term of~\cref{eqn:delta_dot} has the effect of increasing the proportional gain of the control system, since the time derivative of the attitude error vector, \( \dot{e}_{R} \), is linear with respect to the angular velocity error vector \( e_\Omega\).

@@

\subsection{Attitude Parameterizations}\label{ssec:attitude_parameterization}
\end{multicols}
\begin{figure} 
  \centering 
    \subfigure[Configuration error \( \Psi \)\label{fig:Psi_adapt}]{\includegraphics[width=0.4\columnwidth]{Psi_adapt} }~
    \subfigure[Angle to each constraint \label{fig:con_angles}]{\includegraphics[width=0.4\columnwidth]{constraint_angles} }\\
    \subfigure[Disturbance estimate \( \bar \Delta \) components \label{fig:delta_adapt} ]{\includegraphics[width=0.4\columnwidth]{Delta_adapt}}~
    \subfigure[Attitude trajectory \label{fig:cad_adapt}]{\includegraphics[trim={10cm 0 10cm 0},clip,width=0.4\columnwidth]{cad_adapt} }
  \caption{Attitude stabilization with adaptive update law}
  \label{fig:adapt} 
\end{figure}
\begin{multicols}{2}
% Discuss the benefits of using this geometric representation as compared to Euler angles or quaternions
Attitude parameterizations, such as Euler angles and Quaternions, are frequently used in the aerospace and astrodynamics communities~\cite{vallado2007}.
For example, Euler angle sequences are frequently used to describe the transformation between a variety of reference frames used to describe the position and orientation of the orbit of Earth satellites~\cite{vallado2007}.
In addition, quaternions were used during the operation of Skylab and the NASA Space Shuttle~\cite{hughes2004}.
However, the choice of attitude parameterization plays a critical role in control design and the resulting motion of the system.

% singularities and the problems involved with using Euler angles
Euler angle sequences are a minimum, three-parameter set of angles which describe the transformation between two reference frames.
Using Euler angles, we can represent any general rotation as a sequence of three intermediate rotations~\cite{shuster1993}.
By convention, there are \num{24} possible Euler angle sequences for any given rotation.
In addition, Euler angles are a minimum representation, as only three angles, and the associated sequence, are required to describe the three angular degrees of freedom of the rigid body.
However, there is great ambiguity in the representation of the attitude as there are many equivalent Euler angle sequences for a given attitude of the system.
Therefore, great care must be taken in the control system design to ensure that a consistent sequence is used. 
Furthermore, it has been shown that no minimal attitude representation can describe orientations both globally and without singularities~\cite{hughes2004,bhat2000}.
These singularities can cause significant difficulties during control design and hardware implementation.

To demonstrate the effect of the kinematic singularities inherent with Euler angles we will represent the attitude of the body fixed reference frame, \( \vecbf{b}_i \), with respect to the inertial frame, \( \vecbf{e}_i\), in terms of the \( 3-1-3\) Euler angle sequence.
More explicitly, this corresponds to the rotation sequence \( \theta_1 \vecbf{b}_3 , \theta_2 \vecbf{b}_1, \theta_3 \vecbf{b}_3 \).
The rotation matrix, \( R(\theta_1, \theta_2, \theta_3) \), corresponding to this sequence is 
\begin{align}\label{eq:euler313}
    \begin{bmatrix}
        -s_1 c_2 s_3 + c_3 c_1 & -s_1 c_2 c_3 - s_3 c_1 & s_1s_2 \\
        c_1 c_2 s_3 + c_3 s_1 & c_1 c_2 c_3 - s_3 s_1 & - c_1 s_2 \\
        s_2 s_3 & s_2 c_3 & c_2
    \end{bmatrix} ,
\end{align}
where \( s_i, c_i \) represent \( \sin \theta_i, \cos \theta_i \) for \( i = \braces{1,2,3}\).
Using this representation that kinematic differential equations for the associated Euler angles are given as
\begin{align}\label{eq:euler313_diff}
    \begin{bmatrix}
        \dot{\theta}_1 \\ \dot{\theta}_2 \\ \dot{\theta}_3 
    \end{bmatrix}
    =
    \begin{bmatrix}
        \slfrac{\parenth{\Omega_1 s_3 + \Omega_2 c_3}}{s_2} \\
        \Omega_1 c_3 - \Omega_2 s_3 \\
        -\slfrac{\parenth{\Omega_1 s_3 + \omega_2 c_3}c_2}{s_2} + \Omega_3
    \end{bmatrix} .
\end{align}
From~\cref{eq:euler313_diff}, it is immediately clear that a singularity exists when \( \sin \theta_2 = 0 \) or equivalently, \( \theta_2 = 0, \pm \pi \). 
In the vicinity of the singularity, the angular velocities of the Euler angles will tend to approach \( \pm \infty \) and the angular velocities will experience instantaneous sign changes.
Furthermore, all Euler angle sequences will exhibit a similar singularity at either \( \theta_2 = 0, \pm \pi \) or \( \theta_2 = \pm \frac{\pi}{2}, \pm \frac{3\pi}{2} \).
Therefore simply switching the sequence does not alleviate the issue, but rather only moves the singularity.
As a result, Euler angles are not appropriate for systems which experience large angular rotations, such as those demonstrated in~\cref{fig:adapt}, or control systems which rely on the angular velocities \( \theta_i \).
% discuss the unwinding phenomenon with quaternions

\end{multicols}
\begin{figure}[H]
    \centering 
    \subfigure[{Configuration error \( \Psi \)} \label{fig:Psi_tv} ]{\includegraphics[width=0.3\columnwidth]{Psi_tv} }~
    \subfigure[{Angle to constraint } \label{fig:con_angle_tv}]{\includegraphics[width=0.3\columnwidth]{figures/constraint_angle_tv.pdf} }~
    \subfigure[{Disturbance estimate \(\bar\Delta\) components} \label{fig:Delta_tv}]{\includegraphics[width=0.3\columnwidth]{figures/delta_tv} }
    \caption{Time-varying external disturbance simulation}
    \label{fig:tv} 
\end{figure}
\begin{multicols}{2}
\subsection{Time-varying Disturbance}\label{ssec:time_varying}

@@

\Cref{fig:tv} demonstrates the ability for the adaptive controller, which is presented in~\Cref{prop:adaptive_control}, to handle time-varying disturbances.
\Cref{fig:Psi_tv} shows the non-dimensional value of the configuration error function and demonstrates that the adaptive controller is able to stabilize the system to the desired attitude configuration.
In addition,~\cref{fig:con_angle_tv} shows that the constraint is never violated as the angle between the body-fixed sensor \( r \) and the constraint \( v \) is greater than \SI{12}{\degree} over the entire attitude maneuver.
We can see in~\cref{fig:Delta_tv} that that estimate \(\bar \Delta \) for each of the components accurately tracks the true disturbance after approximately~\SI{5}{\second}.

\section{Experiment on Hexrotor UAV}\label{sec:experiment}
\begin{figure}[H]
    \centering
    \includegraphics[width = 0.75\columnwidth]{hexrotor}
    \caption{Attitude control testbed~\label{fig:hexrotor}}
\end{figure}
A hexrotor unmanned aerial vehicle (UAV), as seen in~\cref{fig:hexrotor}, has been developed at the Flight Dynamics and Controls Laboratory (FDCL) at the George Washington University~\cite{kaufman2014}.
The UAV is composed of three pairs of counter-rotating propellers. 
Typical UAVs are composed of four or more co-planar propellers.
As a result, these systems are underactuated and unable to impart a force along every degree of freedom.
For example, quadrotor UAVs are unable to translate laterally without first conducting a rotation.
Conversely, the propeller pairs of the hexrotor are angled relative to one another to allow for a fully actuated rigid body.
This allows the hexrotor to impart a force in any direction and a moment about any axis.
